---
title: "Analysis of RNA-seq data"
author: "Elise Røttereng"
date: "2025-09-22"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Downloading packages

```{r library}
library(tidyverse)
library(DESeq2)
library(gggenes)
library(readxl)
library(tibble)
library(stringr)
library(eulerr)
library(ggplot2)
library(mixOmics)
library(biomaRt)
library(topGO)
library(patchwork)
```

```{r prioritizing functions}
select <- dplyr::select
slice <- dplyr::slice
```

# Read in data


## Data from earlier experiment

```{r data from earlier experiment}

eqtl <- read_tsv(file = "eQTLresult.tsv")

hubtargets <- eqtl %>% 
  filter(eQTL.type=='trans', SNP=='ssa14_50815335') %>% # only trans eQTL and only the gene we are interested in
          distinct(eQTL, .keep_all = T)

head(hubtargets)

```

This is a list of DEGs from an earlier experiment. This data is used for comparison. 

### Genes in the hub-locus

```{r genes in the hublocus}

hub_locus <- c("ENSSSAG00000005903", "ENSSSAG00000052122", "ENSSSAG00000099904", "ENSSSAG00000004575","ENSSSAG00000004590", "ENSSSAG00000004608", "ENSSSAG00000007619")

```

## Core photoperiod genes

```{r  core photoperiod genes}

core_photoperiod <- read_excel("core photoperiod tables2.xlsx")

```

## Read in data from this experiment

The data has been through a nf-core/RNA-seq pipeline

```{r read in data from this experiment}

cts <- read_tsv(file = "rsem.merged.gene_counts.tsv")
head(cts)

# Read in metadata
coldata <- read_excel("bare metadata.xlsx")
head(coldata)

```

## Read in Ensamble gene description

Table over gene id, gene name, description and more downloaded from salmonbase.org at the 19. of september 2025

```{r read in Ensamble gene description}

gene_des <- read.delim("Ensembl_genes.tsv", sep = "\t")
colnames(gene_des)[5] <- "Gene"

```

# Adjusting the data

## cts

```{r adjusting cts}

# Remove transcript_id column
cts <- cts %>% select(-2)

# Make gene_id into row name
cts <- column_to_rownames(cts, var = colnames(cts)[1])

# Counts into "integer"
cts <- round(cts) # round off
cts_matrix <- as.matrix(cts) # make into matrix to be able to change into integer
storage.mode(cts_matrix) <- "integer"

head(cts_matrix)

```

## coldata

```{r adjusting coldata}

# Transpose 
coldata <- t(coldata)

# Tidy
coldata_df <- as.data.frame(coldata)
names(coldata_df) <- coldata_df[1,] # make first row into column name
coldata_df <- coldata_df[-1, , drop = FALSE] # remove first repetitive row

head(coldata_df)

```

## Test if data is in the right format

```{r test if data is in right format}
all(colnames(cts) %in% rownames(coldata))
```

It is correct.

# Run DESeq analysis with photoperiod as design (controll)

This is a quality control (QC).

## Make a DESeq dataset object with photoperiod as design (controll)

```{r make DESeq dataset object with photoperiod}
dds_pp <- DESeqDataSetFromMatrix(countData = cts_matrix,
                       colData = coldata_df,
                       design = ~ light)

dds_pp

# Remove genes with small counts
keep_pp <- rowSums(counts(dds_pp)) >= 10 # All genes get the value True or False after if they have higher counts than 9
dds_pp <- dds_pp[keep_pp,] # Remove False
dds_pp

# Set factor level
dds_pp$light <- relevel(dds_pp$light, ref = "LL") # Set to compare with LL (continous light)
dds_pp$light # Se that LL i first level

```

## Run DESeq2 with photoperiod

```{r run DESeq2 with photoperiod}

dds_des_pp <- DESeq(dds_pp)

# Save the results
res_pp <- results(dds_des_pp)

head(res_pp)
summary(res_pp)
```

## Finding DEGs with photoperiod

Trying to see if the DESeq2 analysis is working by checking if the genes known to be affected by light period are identified. 

### Name of top DGEs with photoperiod

```{r combining data}

# Making a tibbel where the old data and new one is combined
reshub_pp <- res_pp %>% # put the result in reshub
  as_tibble() %>% # make into tibble 
  mutate(Gene=rownames(res_pp)) %>% # make a new column called Gene, same as in hubtarget
  full_join(hubtargets, by = "Gene") # join the together the data by Gene

```

Using padj = 0.05.

```{r name of DEGs}
DEGs_pp <- reshub_pp %>% 
  filter(padj < 0.05) %>% 
#  filter(abs(log2FoldChange) >1) %>% 
  select(padj, Gene, log2FoldChange) %>% 
  merge(select(gene_des, Gene, gene_name, biotype, description), by = "Gene") %>% 
  arrange(padj)

head(DEGs_pp)
```

ENSSSAG00000073055 - ELOVL fatty acid elongase 6 (elovl6)
ENSSSAG00000049592 - same as bellow
ENSSSAG00000005879 - hematopoietic cell-specific Lyn substrate 1(hcls1)
ENSSSAG00000040001 - major histocompatibility complex class I-related gene protein (NCBI)
ENSSSAG00000070750 - protein coding, but unknown function, kind of similar to galectin-9B-like in brown trout
ENSSSAG00000093085 - protein coding, but unknown function, kind of similar to sodium/potassium-transporting ATPase subunit alpha-3-like in brown trout


# Core photoperiod genes


```{r core photoperiod}

# Sanity check 

# Are these known photoperiod genes present in the DEGs

DEGs_pp %>% filter(grepl('sodium/potassium', description)) %>% filter(padj<0.1) 

DEGs_pp %>% filter(grepl('egl-9 family hypoxia-inducible factor 2', description)) %>% filter(padj<0.1)

# The core photoperiod genes
  
sum(core_photoperiod$geneID %in% DEGs_pp$Gene)

DEGs_pp_common <- DEGs_pp %>%
  filter(Gene %in% core_photoperiod$geneID)

```


## Resampling

```{r resampling photo}

set.seed(6)

# Resampling:

# Choose 219 random genes from 41 770 genes, is the overlap with core_photo of 103 a lot?

no_core <- nrow(core_photoperiod)


resampling_overlap_photo <- sapply(1:1000, function(i){
    sum(reshub_pp$Gene[sample(nrow(reshub_pp), no_core, replace = F)] %in% DEGs_pp$Gene)
  })


resampling_overlap_photo #Drawn random we still get a overlap of around 103 genes

median(resampling_overlap_photo)
mean(resampling_overlap_photo)

# calculate the actual overlap 

no_overlap_photo <- sum(core_photoperiod$geneID %in% DEGs_pp$Gene)

no_overlap_photo

# How many times are the random overlap bigger or the same size as the actual one? Divied on the amount of interations. 
sum(resampling_overlap_photo >= no_overlap_photo)/1000 #p-value

```


# DESeq analysis with genotype

## Make a DESeq object

```{r DESeq object with genotype}
dds <- DESeqDataSetFromMatrix(countData = cts_matrix,
                       colData = coldata_df,
                       design = ~ light + genotype) #genotype at the end since this is the one we are most interested in

# Remove genes with low count, less than 10

keep_dds <- rowSums(counts(dds)) >= 10
dds <- dds[keep_dds]

# Set factor level

dds$genotype <- relevel(dds$genotype, ref = "CC")
dds$genotype #confirm that CC is first level

```

## Run DESeq analysis

```{r Run DESeq}

dds_des <- DESeq(dds)

res <- results(dds_des)
res

summary(res)
```

Combining data 

```{r combining new and old data}

# Making a tibbel where the old data and new one is combined
reshub <- res %>% #put the result in reshub
  as_tibble() %>% #make into tibble 
  mutate(Gene=rownames(res)) %>% #make a new column called Gene, same as in hubtargets
  full_join(hubtargets, by = "Gene") #join the together the data by Gene

```

## Expression of the hublocus genes and their genotype

```{r boxplot expression hublocus genes}

set.seed(14)

genot <- as.data.frame(coldata) %>% slice(-1)
cts_t <- as_tibble(t(cts))
cts_t <- cts_t %>% mutate(ID = colnames(cts))
cts_t$genot <- genot$V1[match(cts_t$ID, rownames(genot))]

annot <- read_tsv('https://salmobase.org/datafiles/TSV/genes/AtlanticSalmon/Ssal_v3.1/Ensembl_genes.tsv')

gene_map <- annot %>% 
  select(gene_id, gene_name) %>% 
  distinct()


fill_cols <- c("CTSZ CC"= '#F47F73',
              "CTSZ TT"= alpha('#F47F73', 0.5),
              "MPPE1 CC"= '#FBB463', 
              "MPPE1 TT"= alpha('#FBB463', 0.5),
              "GNAI1 CC"= '#CDE6C4', 
              "GNAI1 TT"= alpha('#CDE6C4', 0.5),
              "pak1Ip1 CC"= "#8ED2C6",
              "pak1Ip1 TT"= alpha("#8ED2C6", 0.5),
              "endoU CC"='#FCF6B5', 
              "endoU TT"=alpha('#FCF6B5', 0.5),
              "FAM111A CC"='#BFBADA',
              "FAM111A TT"= alpha('#BFBADA', 0.5),
              "ZFYVE9 CC"="#81B2D3", 
              "ZFYVE9 TT"=alpha("#81B2D3", 0.5))

gene_manual <- c(
  "ENSSSAG00000005903" = "CTSZ",
  "ENSSSAG00000052122" = "MPPE1",
  "ENSSSAG00000099904" = "GNAI1",
  "ENSSSAG00000004575" = "pak1Ip1",
  "ENSSSAG00000004590" = "endoU",
  "ENSSSAG00000004608" = "FAM111A",
  "ENSSSAG00000007619" = "ZFYVE9"
)

cts_t_long <- cts_t %>% 
  select(hub_locus, genot) %>% 
  pivot_longer(cols = starts_with("ENS"), values_to = 'counts') %>% 
  dplyr::rename(genes = name) %>% 
  left_join(gene_map, by = c("genes" = "gene_id")) %>% 
  mutate(
    gene_name = dplyr::coalesce(gene_name, gene_manual[genes]),
    gene_name = ifelse(is.na(gene_name), genes, gene_name),
    counts = log2(counts + 1),
    group = factor(paste(gene_name, genot), levels = unlist(lapply(gene_manual, function(g) paste(g, c("CC","TT"))))))

p0 <- cts_t_long %>% 
  ggplot(aes(y = counts, x = group, fill = group)) +
  geom_boxplot() +
  scale_fill_manual(values = fill_cols) +
  geom_jitter(aes(color = group), shape = 21, size = 2.5, color = "black") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1), legend.position = "none") +
  plot_layout(widths = 555, heights = 550) +
  labs(x = "Gene and genotype", y = "log2(BaseMean + 1)")


fill_colors <- c(
  "CTSZ" = '#F47F73',
  "MPPE1" = '#FBB463',
  "GNAI1" = '#CDE6C4',
  "pak1Ip1" = "#8ED2C6",
  "endoU" = '#FCF6B5',
  "FAM111A" = '#BFBADA',
  "ZFYVE9" = "#81B2D3")

p1 <- annot %>% 
  filter(seqname == '14', start > 50400000 & end < 51000000, gene_id %in% hub_locus) %>%
  mutate(
    gene_name = dplyr::coalesce(gene_manual[gene_id], gene_id),
    gene_name = factor(gene_name, levels = names(fill_colors)),
    strand = strand == '+') %>% 
  ggplot(aes(xmin = start, xmax = end, y = seqname, fill = gene_name,forward = strand)) +
  geom_gene_arrow() +
  scale_fill_manual(values = fill_colors) +
  xlim(c(50400000, 51000000)) +
  theme_genes() +
  labs(x = "Position (Mbp)", y = "", fill = "Gene")
 
 
patch_hublocus_expr_plot <- p1/p0 + plot_annotation(title = "Gene expression of the hub-locus genes", tag_levels = "A")


patch_hublocus_expr_plot

```

## Gen expression of the hub-locus genes

```{r gen expr hub-locus}

locus_in_DEGs <- reshub %>% 
  filter(Gene %in% hub_locus)

fill_cols_2 <- c("ENSSSAG00000005903"= '#F47F73',
              "ENSSSAG00000052122"= '#FBB463',
              "ENSSSAG00000099904"= '#CDE6C4',
              "ENSSSAG00000004575"= "#8ED2C6",
              "ENSSSAG00000004608"='#BFBADA',
              "ENSSSAG00000004590"='#FCF6B5', 
              "ENSSSAG00000007619"= '#81B2D3')


locus_in_DEGs <- locus_in_DEGs %>%
  mutate(Gene = factor(Gene, levels = names(fill_cols_2)))

ggplot(locus_in_DEGs, aes(x = Gene, y = baseMean, fill = Gene)) +
  geom_col() +
  scale_fill_manual(values = fill_cols_2) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Gene Expression of the Genes in the Hub-Locus")

```

# DEGs

## Amout of DEGs

Using padj = 0.2.

```{r amount of DEGs}

# How many DEGs are found in the hubsomt data?
reshub <- reshub %>% 
  mutate(is.hub=ifelse(is.na(eQTL), F, T)) # make is.hub with True or False if eQTL is na, if Na = False gene is found in hubtargets

reshub %>% filter(padj<0.2) %>% # filter for adjusted p-value = 0.2
  group_by(is.hub) %>% 
  summarize(mean.lfc = mean(log2FoldChange), N = n()) # find mean for log2FoldChange and count how many genes are NA

```

440 Degs only in the new data, only 24 of them are also found in the hubtarget data. Low mean log2 fold change.

## Percentage of DEGs in the differnt data sets

```{r DEG %}
440/(41780-2075) # number of DEGs only in the new data divided by number of genes only in the new data

24/2075 # number of DEGs in both data sets divided by number of DEGs form the hubtarget data
```

## Name of top DEGs overall

```{r top DEGs}
DEGs <- reshub %>% 
  filter(padj < 0.2) %>% 
  select(padj, Gene, log2FoldChange, baseMean) %>% 
  merge(select(gene_des, Gene, gene_name, biotype, description), by = "Gene") %>% 
  arrange(padj)

head(DEGs)
```

ENSSSAG00000004590 - uridylate-specific endoribonuclease C-like (NCBI)
ENSSSAG00000007619 - protein coding, but unknown function, kind of similar to zinc finger FYVE domain-containing protein 9-like in brown trout
ENSSSAG00000113063 - protein coding, but unknown function
ENSSSAG00000004608 - protein coding, but unknown function
ENSSSAG00000110711 - long non-coding RNA
ENSSSAG00000118211 - protein coding, but unknown function

## Name of top DEGs that are also found in Hubtarget

```{r name of DEGs found in both data sets}
DEGs_hubtargets <- reshub %>%
  filter(is.hub == TRUE, padj < 0.2) %>% # only include genes that are in both data set and have an adjusted p-value over 0.2
  select(padj, Gene, log2FoldChange, baseMean) %>% 
  merge(select(gene_des, Gene, gene_name, biotype, description), by = "Gene") %>% 
  arrange(padj)

head(DEGs_hubtargets)
```

ENSSSAG00000040831 - CCR4-NOT transcription complex, subunit 10 (cnot10)
ENSSSAG00000005545 - KRIT1, ankyrin repeat containing (NCBI)
ENSSSAG00000041952 - transmembrane protein 128 (tmem107)
ENSSSAG00000063494 - solute carrier family 9 member A6 (slc9a6a)
ENSSSAG00000084204 - protein coding, but unknown function
ENSSSAG00000000827 - peroxiredoxin 4 (prdx4)


## Venn diagram over DEGs

```{r Venn diagram}

DEGs_names <- DEGs$Gene
Hubtargets_names <- hubtargets$Gene

gene_name_list <- list("New DEGs" = DEGs_names, "Hubtargets" = Hubtargets_names)

fit <- euler(gene_name_list)

venn_plot <- plot(fit,
     quantities = T,
     fills = c("steelblue", "deeppink4"),
     alpha = 0.5,
     labels = TRUE,
     main = "The amount of genes overlapping between this experiment and the hubtargets")

venn_plot

```


# Statstical control 

Want to find how statistically strong the DEGs are. 

## Randomization test

```{r randomization test hubtargets}

set.seed(14)

# Draw DEGS amount of random genes
no_DEGs <- nrow(DEGs) 

# Making a table with expression data and a row with T/F after if it is a DEG and if it is in hubtargets. Not used directly in the randomization test
expr_tbl <- reshub %>% select(Gene, baseMean, log2FoldChange, pvalue, padj) %>% 
  filter(!is.na(baseMean)) %>% 
  mutate(is_DEG = Gene %in% DEGs$Gene) %>% 
  mutate(is_hub = Gene %in% hubtargets$Gene)
  

# Resampling:

# Choose 464 random genes from 41 770 genes, is the overlap with hubtagets of 24 a lot?
resampling_overlap <- sapply(1:1000, function(i){
    sum(reshub$Gene[sample(nrow(reshub), no_DEGs, replace = F)] %in% hubtargets$Gene)
  })

resampling_overlap #Drawn random we still get a overlap of around 24 genes

median(resampling_overlap)
mean(resampling_overlap)

# calculate the actual overlap 
no_overlap_hub <- DEGs %>% 
  filter(Gene %in% hubtargets$Gene) %>% 
  nrow()

no_overlap_hub

# How many times are the random overlap bigger or the same size as the actual one? Divied on the amount of interations. 
sum(resampling_overlap >= no_overlap_hub)/1000 #p-value

```


## PLS-DA

```{r rlog transformasjon}

rld <- rlog(dds_des, blind = TRUE) #use blind = TRUE so design is ignored and does not impact the test
```

### Run PLS-DA

```{r Run PLS-DA}

# Get the matrix from the rld-object
expr_matrix <- t(assay(rld))  #turn the matrix

# Remove the effect of light
light <- colData(dds_des)$light

expr_matrix_corrected <- apply(expr_matrix, MARGIN = 2, function(gene_expr) { #apply + MARGIN = 2 means run for every column, gene_expr is the expression of every sample
  residuals(lm(gene_expr ~ light)) #run a linear regression on the gen expression and use the light as a explanation variable, residual keeps the gen expression that can not be explained by light. 
})

# Make group vector
group_vector <- colData(dds_des)$genotype

# Run the PLS-DA with 3 components
plsda_result <- plsda(expr_matrix_corrected, group_vector, ncomp = 5)

```


### Cross-validation

```{r cross validation PLS-DA}

set.seed(2002)

# Cross-validation of the PLS_DA model
perf_result <- perf(plsda_result, validation = "Mfold", folds = 5) 

# Plot of error rates
plot(perf_result$error.rate$overall[,"max.dist"], type = "b",
     xlab = "Number of components", ylab = "Error rate",
     main = "Cross-validation Error for PLS-DA",
     xaxt = "n")

axis(1, at = 1:nrow(perf_result$error.rate$overall), labels = rownames(perf_result$error.rate$overall))

```

### Score Plots

```{r Score plots}

# Plot the frst two components against each other
df_comp12 <- data.frame(comp1 = plsda_result$variates$X[,1],
                 comp2 = plsda_result$variates$X[,2],
                 group = group_vector)

PLS_DA_plot<- ggplot(df_comp12, aes(x = comp1, y = comp2, color = group)) + geom_point(size = 3, alpha = 0.8) +
  labs(title = "PLS-DA score plot: Component 1 vs 2", 
       x = "Component 1", 
       y = "Component 2", 
       color = "Genotype") +
  theme_classic()

PLS_DA_plot #every dot is a sample


# Scores for component 1
scores1 <- plsda_result$variates$X[,1] 
scores1_df <- data.frame(scores1)

# Box plot of how well component two is to separate genotypes
score1_plot <- ggplot(scores1_df, aes(x = group_vector, y = scores1, fill = group_vector)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, size = 2, color = "black") +
  labs(title = "Component 1",
       x = "Genotype",
       y = "Component 1 score",
       fill = "Genotype") +
  theme_classic()

score1_plot

# Scores for component 2
scores2 <- plsda_result$variates$X[,2] 
scores2_df <- data.frame(scores2)

# Box plot of how well component two is to separete genotypes
score2_plot <- ggplot(scores2_df, aes(x = group_vector, y = scores2, fill = group_vector)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, size = 2, color = "black") +
  labs(title = "Component 2",
       x = "Genotype",
       y = "Component 2 score",
       fill = "Genotype") +
  theme_classic()

score2_plot

patchwork_score_plots <- score1_plot | score2_plot
patchwork_score_plots + plot_annotation(title = "Separation on PLS-DA", tag_levels = "A")
```

### PLS-DA loadings

```{r PLS-DA loadings}

loadings_plot <- plotLoadings(plsda_result, comp = 1, method = "mean")


load <- plsda_result$loadings$X[, 1]

loadings_plot_2 <- plot(load,
     pch = 16,
     xlab = "Genes index",
     ylab = "Loading value",
     main = "Loadings to Componenet 1")



# Make loadings into dataframe
loadings_DA <- plsda_result$loadings
df_loadings_DA <- as.data.frame(loadings_DA)
head(df_loadings_DA)

# Top 6 genes with higest absolut loading
top6_genes_comp1_DA <- df_loadings_DA %>%
  rownames_to_column("Gene") %>%
  select(Gene, Loading = X.comp1) %>% 
  inner_join(select(gene_des, Gene, gene_name, biotype, description), by = "Gene") %>% arrange(desc(abs(Loading))) %>%
  head()

top6_genes_comp1_DA

```

### Calculating VIP-scores

```{r calculating VIP scores}

vip_scores_DA <- mixOmics::vip(plsda_result)
head(vip_scores_DA)

# Calculating VIP-scores combined for the components
vip_classic <- function(plsda_obj) {
  W <- plsda_obj$loadings$X        #w_jk (variables x components)
  T <- plsda_obj$variates$X        #scores (observations x components)
  Q <- plsda_obj$loadings$Y        #y-loadings (dependency of y)
  
  p <- nrow(W)  #number of variables
  A <- ncol(W)  #number of components

  # Calculate SSY for each componenet
  SSY <- colSums((T %*% t(Q))^2)
  
  # Classic VIP
  vip <- rep(0, p)
  for (j in 1:p) {
    vip[j] <- sqrt(p * sum(SSY * (W[j, ]^2)) / sum(SSY))
  }
  
  names(vip) <- rownames(W)
  return(vip)
}


vip_final <- vip_classic(plsda_result)

vip_sorted <- sort(vip_final, decreasing = TRUE) %>%
  enframe(name = "Gene", value = "value") %>%
  inner_join(select(gene_des, Gene, gene_name, biotype, description), by = "Gene") %>%
  as_tibble()

head(vip_sorted)

top30VIP <- slice_head(vip_sorted, n = 30)

top_30_vip_plot <- ggplot(top30VIP, aes(x = Gene, y = value)) +
  geom_col(fill = "salmon") +
  coord_flip() +
  labs(title = "The Top 30 Genes with the Highest VIP Scores",
       x = "Genes",
       y = "VIP score") +
  theme_minimal()

top_30_vip_plot

```

How many of the genes with high VIP score are hubtargets

```{r hutargets in top VIP}

hubtargets_in_VIP <- top30VIP %>% 
  filter(Gene %in% hubtargets$Gene)

nrow(hubtargets_in_VIP)

hubtargets_in_VIP

```

How many of the genes with high VIP are in the hub-locus

```{r vip to locus}

locus_in_VIP <- top30VIP %>% 
  filter(Gene %in% hub_locus)

nrow(locus_in_VIP)

locus_in_VIP

```

# Pathway analysis

## Gen ontology (GO) enrichment anlysis

### Seting up for running GO

```{r GO preparation}

# Filtering for log2FoldChange
DEGs_fold <- subset(DEGs, abs(log2FoldChange) > 0.5) #from 464 genes to only 141 genes
DEGs_list <- DEGs_fold$Gene #making list of gene names


# Conecting to Ensamble for salmon 
ensembl <- useMart("ensembl", dataset = "ssalar_gene_ensembl")


# Annotating all genes
annot <- getBM(
  attributes = c("ensembl_gene_id", "go_id"), #get ensembl-ID and GO-ID
  mart = ensembl) %>% #gens are named in teh ensembl format
  filter(go_id != "") #filer out the genes that do not have an ensembl ID


# List of Ensembl ID with GO-ID
gene2GO <- split(annot$go_id, annot$ensembl_gene_id) #put the GO-IDs together with each gene 


# List over genes and if they are DEGs or not
gene_list <- annot %>% 
  distinct(ensembl_gene_id) %>% #every genes is only in one time
  mutate(isDEG = as.integer(ensembl_gene_id %in% DEGs_fold$Gene)) %>% #make a new colunm with 1 or 0 if it is a DEG or not
  deframe() %>% #makes into a vector
  factor() #convert 0 and 1 into factor

```

### Running GO-enrichment for biological process

```{r run GO BP}

# Create topGOdata object 
GOdata_BP <- new("topGOdata",
              ontology = "BP",  #Biological Process
              allGenes = gene_list,
              annot = annFUN.gene2GO,
              gene2GO = gene2GO)


# See if GO-terms is more common in DEGs then is expected
res_fisher_BP <- runTest(GOdata_BP, algorithm = "classic", statistic = "fisher")
res_fisher_BP

# 223 non-trivial GO-terms, terms with at least one DEG. 21 GO-terms had significant over representations of DEGs. Only 44 DEGs with GO-information 

# Table with the 20 most significant GO-terms (gene functions)
GOres_BP <- GenTable(GOdata_BP, classicFisher = res_fisher_BP, topNodes = 20)
head(GOres_BP)

# Clean up
GOres_clean_BP <- GOres_BP %>%
  mutate(
    classicFisher = as.numeric(classicFisher), #makes the p-values into numbers from character 
    negLog10P = -log10(classicFisher)) %>% #logfold the p-value to make it readable, minus so lower numbers get higher value
  filter(!is.na(classicFisher)) %>% #remove rows without p-value
  arrange(desc(negLog10P)) %>% #most significsnt first
  mutate(Term = factor(Term, levels = rev(unique(Term))))

# Make plot
GO_plot_BP <- ggplot(GOres_clean_BP[1:20, ], aes(x = Term, y = negLog10P)) +
  geom_col(fill = "#0072B2") +
  coord_flip() +
  ylim(0, 5.5) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Biological Process",
    x = NULL,
    y = expression(-log[10](p-value)))

GO_plot_BP

```

### Run GO-enrichment for molecular function

```{r run GO MF}

# Run GO-enrichment
GOdata_MF <- new("topGOdata",
              ontology = "MF",  #Molecular function
              allGenes = gene_list,
              annot = annFUN.gene2GO,
              gene2GO = gene2GO)


# See if GO-terms is more common in DEGs then is expected
resultFisher_MF <- runTest(GOdata_MF, algorithm = "classic", statistic = "fisher")
resultFisher_MF

# 208 non-trivial GO-terms, terms with at least one DEG. 10 GO-terms had significant over representations of DEGs. Only 94 DEGs with GO-information 

# Table with the 20 most significant GO-terms (gene functions)
GOres_MF <- GenTable(GOdata_MF, classicFisher = resultFisher_MF, topNodes = 20)
head(GOres_MF)

# Clean up
GOres_clean_MF <- GOres_MF %>%
  mutate(
    classicFisher = as.numeric(classicFisher), #makes the p-values into numbers from character 
    negLog10P = -log10(classicFisher)) %>% #logfold the p-value to make it readable, minus so lower numbers get higher value
  filter(!is.na(classicFisher)) %>% #remove rows without p-value
  arrange(desc(negLog10P)) %>% #most significsnt first
  mutate(Term = factor(Term, levels = rev(unique(Term))))

# Make plot
GO_plot_MF <- ggplot(GOres_clean_MF[1:20, ], aes(x = Term, y = negLog10P)) +
  geom_col(fill = "#EE3A8C") +
  coord_flip() +
  ylim(0, 5.5) +
  theme_minimal(base_size = 13) +
  labs(
    title = "Molecular function",
    x = NULL,
    y = expression(-log[10](p-vaue)))

GO_plot_MF

```

Patchwork plot

```{r patchwork GO plot}
patch_GO_plot <- GO_plot_BP | GO_plot_MF
patch_GO_plot + plot_annotation(
  title = "Significant GO-terms",
  tag_levels = "A")
```


# Compare to scRNA

```{r load}

load("all_couts_genes_object.Rdata")

scRNA_matrix <- sum_matrix

rm(sum_matrix)

```

Give same name

```{r name}

sample_map <- c(
  sample1 = "GL217",
  sample2 = "GL236",
  sample3 = "GL218",
  sample5 = "GL225",
  sample7 = "GL226",
  sample9 = "GL233",
  sample10 = "GL235",
  sample11 = "GL219",
  sample12 = "GL234")

```

Keep only the common genes

```{r common genes}

common_genes <- intersect(rownames(scRNA_matrix), rownames(cts_matrix))

sc_sub  <- scRNA_matrix[common_genes, ]
cts_sub  <- cts_matrix[common_genes, ]

```

Use only the samples that are in both scRNA and bulkRNA

```{r common samples}

sc_samples <- intersect(colnames(sc_sub), names(sample_map))
bulk_samples <- sample_map[sc_samples]

# Get mean of expression
sc_agg <- rowMeans(scRNA_matrix[, sc_samples])

bulk_agg <- rowMeans(cts_matrix[, bulk_samples])
```

Remove genes with 0 expression

```{r remove zero}

keep <- sc_agg > 0 | bulk_agg > 0
sc_agg <- sc_agg[keep]
bulk_agg <- bulk_agg[keep]

```

Plot

```{r plot}

plot(
  log10(bulk_agg),
  log10(sc_agg),
  pch = 16,
  cex = 0.4,
  xlab = "Bulk RNA (log10)",
  ylab = "scRNA pseudo-bulk (log10)",
  main = "All samples aggregated")


```

